---
title: "Change in Homicide Rates (2022 vs. 2019-2021 Avg)"
author: "Ethan Covert"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This R script represents the homicide rates in U.S. states. The rates displayed on the map is the difference from 2022 to the average across 2019 to 2021. Everything mapped is either a steady rate, increased rate, or decreased rate - magnitude displayed by color and arrows. 
The data was found here: <https://www.cdc.gov/nchs/pressroom/sosmap/homicide_mortality/homicide.htm>.

## Graph
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=18, fig.height=10, out.width="100%"}
library(usmap)       # U.S. map visualization
library(usmapdata)   # State centroid labels
library(ggplot2)     # Data visualization
library(tidyverse)   # Data manipulation
library(sf)          # Handling spatial data
library(gridExtra)   # For arranging plots and tables together

# Load the dataset
data <- read_csv("homicide_rates.csv")

# Convert 'RATE' to numeric, replacing NAs with 0
data$RATE <- as.numeric(data$RATE)
data$RATE[is.na(data$RATE)] <- 0

# Calculate the average RATE from 2019 to 2021
avg_rate <- data %>%
  filter(YEAR >= 2019 & YEAR <= 2021) %>%
  group_by(STATE) %>%
  summarise(AVG_RATE_2019_2021 = mean(RATE, na.rm = TRUE))

# Extract 2022 rates
rate_2022 <- data %>%
  filter(YEAR == 2022) %>%
  select(STATE, RATE_2022 = RATE)

# Merge the data frames and calculate the difference
final_data <- merge(avg_rate, rate_2022, by = "STATE", all = TRUE)
final_data$RATE_DIFF <- final_data$RATE_2022 - final_data$AVG_RATE_2019_2021

# Replace any remaining NA values with 0
final_data[is.na(final_data)] <- 0

# Categorize rate changes into discrete tiers with arrows
final_data <- final_data %>%
  mutate(
    ARROW = case_when(
      RATE_DIFF == 0 ~ "-",
      RATE_DIFF > 0 & RATE_DIFF <= 1 ~ "↑",
      RATE_DIFF > 1 & RATE_DIFF <= 2 ~ "↑↑",
      RATE_DIFF > 2 ~ "↑↑↑",
      RATE_DIFF < 0 & RATE_DIFF >= -1 ~ "↓",
      RATE_DIFF < -1 & RATE_DIFF >= -2 ~ "↓↓",
      RATE_DIFF < -2 ~ "↓↓↓"
    ),
    LABEL = paste(STATE, ARROW)  # Combine state abbreviation with arrows
  )

# List of small states that have their own table (added Hawaii)
small_states <- c("VT", "NH", "CT", "RI", "NJ", "DE", "MD", "MA", "ME", "HI")

# Create a table of small states with their labels
small_states_table <- final_data %>%
  filter(STATE %in% small_states) %>%
  select(STATE, ARROW) %>%
  arrange(STATE)

# Remove small states from the map dataset (but keep their colors)
final_data <- final_data %>%
  mutate(LABEL = ifelse(STATE %in% small_states, "", LABEL))  # Remove labels for small states

# Convert state abbreviations to full state names for mapping
final_data <- final_data %>%
  mutate(STATE_FULL = state.name[match(STATE, state.abb)])

# Remove any NA values (from mismatches)
final_data <- na.omit(final_data)

# Get centroid coordinates for states
centroid_labels <- usmapdata::centroid_labels("states")

# Extract numeric coordinates from `geom` column
centroid_labels$x <- st_coordinates(centroid_labels$geom)[,1]  # Longitude
centroid_labels$y <- st_coordinates(centroid_labels$geom)[,2]  # Latitude

# Convert state names to lowercase for merging
final_data <- final_data %>%
  mutate(STATE_FULL = tolower(STATE_FULL))

centroid_labels <- centroid_labels %>%
  mutate(STATE_FULL = tolower(full))

# Merge homicide data with centroid coordinates
final_data <- final_data %>%
  inner_join(centroid_labels, by = "STATE_FULL")

# Create a legend manually with both numbers and arrows
legend_labels <- c("-3 ↓↓↓", "-2 ↓↓", "-1 ↓", " 0 -", " 1 ↑", " 2 ↑↑", " 3 ↑↑↑")

# Create the US Map plot with colors for all states
map_plot <- plot_usmap(data = final_data, values = "RATE_DIFF", regions = "states") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, 
                       name = "Rate Change", breaks = c(-3, -2, -1, 0, 1, 2, 3),
                       labels = legend_labels) +
  labs(title = "Change in Homicide Rates (2022 vs. 2019-2021 Avg)",
       subtitle = "Red = Increase, Blue = Decrease. Arrows indicate magnitude.") +

  # 🔥 **Ensure the theme does not remove the map details**
  theme_minimal() +  # Keeps map elements visible while allowing text adjustments
  theme(
    plot.title = element_text(size = 24, face = "bold"),   # Title size
    plot.subtitle = element_text(size = 18),              # Subtitle size
    legend.title = element_text(size = 20, face = "bold"), # Legend title size
    legend.text = element_text(size = 16),                # Legend labels size
    legend.key.size = unit(1.5, "cm")                     # Makes legend boxes bigger
  ) +

  # 🔥 **Re-add state labels after theme adjustments**
  geom_text(data = final_data, aes(x = x, y = y, label = LABEL), 
            size = 4, color = "black", fontface = "bold")  # Proper centroid alignment

# Create a table of small states without rate difference
table_plot <- tableGrob(small_states_table, rows = NULL)

# Arrange map and table side by side
grid.arrange(map_plot, table_plot, ncol = 2, widths = c(3, 1))
```

## Observations

Some observations can be made from this data following the trends. States like California or Texas are bigger states while having very little change. On the mention of bigger states, Alaska has the largest increase despite its rural geography. On the other hand, Wyoming has the greatest decrease despite also having a very rural geography. Alaska has had a relatively constant homicide rate, recently shooting into the higher end, while Wyoming had a low rate, and has now dropped heavily. Alaska's crime rate can be related to recent increases in drug and violent crime. For example, in 2022 26.85 Kilograms of fentanyl were seized in Alaska (Alaska Department of Public Safety, 2022b). Rape was on an increase, almost 25% in 2022, along with Child sexual abuse doubling from the previous year (Alaska Department of Public Safety, 2022b). However, we see Wyoming's homicide rates go down as other violent crimes also decreased. Sexual assault decreased around 10%, and almost 90% of sexual assaults involved no weapons (other than hands and feet) - leading to the decrease in homicides (Wyoming Division of Criminal Investigation.). Through investigating each state with a high rate of change, we can begin to piece together what causes what. Looking at Wyoming and Alaska, the two states with the highest rates in each direction, I see that violent crime and sex crimes is heavily related to homicide rates.


Alaska Department of Public Safety. (2022a). 2022 Alaska dashboard.
https://dps.alaska.gov/getmedia/eace8a7b-e279-4d03-a424-3f9129aa9643/2022-Alaska-Dashboard.pdf

Alaska Department of Public Safety. (2022b). Alaska law enforcement seizes millions of doses of fentanyl, methamphetamine, and other illicit drugs. https://dps.alaska.gov/AST/PIO/PressReleases/Alaska-Law-Enforcement-Seizes-Millions-of-Doses-of

Wyoming Division of Criminal Investigation. (2022). Violent crimes in Wyoming, 2022. https://crimestats.wyo.gov/tops/report/violent-crimes/wyoming/2022